---
title: "Analysis of the effect of vitamin C on tooth growth in guinea pigs"
author: "Eduardo Bortoluzzi Junior"
date: "October 20, 2015"
output: pdf_document
---

# Synopsis
<!-- In a few (2-3) sentences explain what is going to be reported on. -->
The data analysed contains the length of the teeth in each of 10 guinea pigs at
each of three dose levels of Vitamin C (0.5, 1 and 2 mg) with each of two
delivery methods (orange juice or ascorbic acid). The growth effect of each
pair of dose level and delivery method is analysed is this report. At least,
with the increase of the dose, the tooth grows more.

# Loading and processing the data

The data with the response of the Vitamin C administration is found in the 
ToothGrowth data frame, which is in the datasets' R package.

```{r}
library(datasets)
library(ggplot2)
```

Let's start exploring the ToothGrowth data frame:

```{r}
str(ToothGrowth)
```

There are 3 variables:

* len: the tooth length
* supp: supplement type:
    * VC: ascorbic acid
    * OJ: orange juice
* dose: dose in milligrams

And there are 60 observations. We can divide the dataframe in 6 groups,
each group with a supplement type and a specific dose. Each group has 10 
observations.

The summary of the data is shown at the plot below:

```{r}
ggplot(data=ToothGrowth, 
       aes(x=dose , y=len, fill=supp, group=interaction(dose,supp))) + 
geom_boxplot() +
ggtitle("Tooth growth in guinea pigs by supplement and dose")
```

Let's test, in each dose, which supplement type effects more in the tooth growth.

<!--
The mean and the standard deviation of each group is needed to do the hypothesis
tests:

```{r}
sample <- with(ToothGrowth, 
               aggregate(len,
                         by=list(dose=dose, supp=supp), 
                         FUN=function(x) c(mean=mean(x), sd=sd(x))
                        )
               )

sample
```
-->
Using the supplement type orange juice as the base, let's make a null hypothesis
$H_0$ being that the growth using the ascorbic acid is the same as the orange
juice, so, $H_0: \mu_{OJ} = \mu_{VC}$. Our alternative hypothesis will be that
the tooth grows less or more when using the ascorbic acid as the supplement, so,
$H_a: \mu_{OJ} \neq \mu_{VC}$
 
For 0.5 mg administration, the confidence interval is:

```{r}
adm <- 0.5
with(ToothGrowth, 
     t.test(ToothGrowth[dose==adm & supp=="VC",]$len,
            ToothGrowth[dose==adm & supp=="OJ",]$len,
            var.equal = FALSE,
            paired = FALSE)$conf.int
)
```

So, as the interval is less than 0, we reject the $H_0$ and accept $H_a$, with
the confidence that the tooth grows less when using the ascorbic acid.

For 1.0 mg administration, the confidence interval is:

```{r}
adm <- 1
with(ToothGrowth, 
     t.test(ToothGrowth[dose==adm & supp=="VC",]$len,
            ToothGrowth[dose==adm & supp=="OJ",]$len,
            var.equal = FALSE,
            paired = FALSE)$conf.int
)
```

The interval is also less than 0, so we also reject $H_0$ in favor of $H_a$. The
tooth grows less when using ascorbic acid.

For 2.0 mg administration, the confidence interval is:

```{r}
adm <- 2
with(ToothGrowth, 
     t.test(ToothGrowth[dose==adm & supp=="VC",]$len,
            ToothGrowth[dose==adm & supp=="OJ",]$len,
            var.equal = FALSE,
            paired = FALSE)$conf.int
)
```

In this case, the interval contains 0, se we fail to reject $H_0$. The tooth 
may grow more or less than using the oranje juice supplement.

For a final hypothesis test, let's give a null hypothesis that the tooth growth
using oranje juice supplement is the same for 1.0 mg and 2.0 mg administration,
so $H_0: \mu_{OJ.1} = \mu_{OJ.2}$. The alternative hypothesis is that they are
different, $H_a: \mu_{OJ.1} \neq \mu_{OJ.2}$.

```{r}
with(ToothGrowth, 
     t.test(ToothGrowth[dose==2.0 & supp=="OJ",]$len,
            ToothGrowth[dose==1.0 & supp=="OJ",]$len,
            var.equal = FALSE,
            paired = FALSE)$conf.int
)
```

The confidence interval does no have 0, and is slightly above it, so the $H_0$
can be reject in favor of $H_a$.

# Results

Vitamin C administered via orange juice causes more tooth growth than using the
acid ascorbic as the supplement, for 0.5 mg and 1.0 mg doses. For 2.0 mg dose,
it is not conclusive, but, due the higher variance when using acid ascorbic as
the supplement, the orange juice can be used when a more controlled discharge
is needed.

A higher tooth growth can be a problem, so it is recommend to use acid ascorbic,
as a Vitamin C supplement, when this collateral effect needs to be minimized.